---
title: "04_DGE_4M_month"
author: "Natalia Andrade Rodriguez"
date: "2023-09-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Gene expresion analysis for all samples that passed quality check and normalization (see 03_Normalizing_All_4M.Rmd)

### Packages
```{r}
# library(DESeq2)
# library(ggplot2)
# library(ggrepel)
# library(dplyr)
# library(tidyr)
# library(reshape2)
# library(tidyverse)
```


### Loading data
```{r}
colData_4M<- readRDS("cache/colData_4M.rds")
data.counts<-readRDS("cache/data_counts_4M.rds")
```

- The annotations for each transcript
```{r}
anno<- readRDS("cache/count_data_35821_anno.rda") %>% mutate(ofav_gene = GeneID)

anno<- anno[,c(164:193)]
```

- making the rownames and column names identical
```{r}
all(rownames(colData_4M) %in% colnames(data.counts))
all(rownames(colData_4M) == colnames(data.counts))
```


### Setting up model design : ~ SCTLD_RNA_Region_season
This model will will allow us to compare affected colonies per sample months ('sep_unaffected vs winter_unaffected' etc). Here we can compare  the regions independently as ECA correspond to fallust and winterch and Lower Keys correspond to fallember and winterruary.

```{r}
dds_sctld.reg.season<- DESeqDataSetFromMatrix(countData = data.counts, 
                            
                                colData = colData_4M, 
 
                                design = ~ 0 + SCTLD_RNA_Region_season )
```

#### Fitting models
```{r}
if (file.exists("cache/dds_sctld.reg.season.rds")) {
  dds_sctld.reg.season <- read_rds("cache/dds_sctld.reg.season.rds")
} else {
 dds_sctld.reg.season <- DESeq(dds_sctld.reg.season, parallel = TRUE)
  write_rds(dds_sctld.reg.season,"cache/dds_sctld.reg.season.rds")
}

beep()
```

#### Transform reads with variance Stabilizing Transformation
```{r}
vst_dds_sctld.reg.season = vst(dds_sctld.reg.season, blind=FALSE)

beep()
```

#### Transform reads with rlog transfroamtion 
```{r}

if (file.exists("cache/rlog_dds_sctld.reg.season.rds")) {
  rlog_dds_sctld.reg.season <- read_rds("cache/rlog_dds_sctld.reg.season.rds")
} else {
rlog_dds_sctld.reg.season <- rlog(dds_sctld.reg.season, blind=FALSE)
  write_rds(rlog_dds_sctld.reg.season,"cache/rlog_dds_sctld.reg.season.rds")
}


beep()
```

### Identify genes with high Cook's distance

#### Cook’s distance plot
```{r}
# ~ SCTLD_RNA_Region_season
# png("results/5_All_4M_DEG/cookdistance_dds_sctld.reg.season.png", width=30, height=8, units = "in", res = 300)
boxplot_dds_sctld.reg.season<-boxplot(log10(assays(dds_sctld.reg.season)[["cooks"]]+1), names = colData_4M$colnames,cex.axis = 0.7, las = 3,range=2,main="Cook's distance ~ SCTLD_RNA_Region_season")
# dev.off()

boxplot_dds_sctld.reg.season<-boxplot(log10(assays(dds_sctld.reg.season)[["cooks"]]), names = colData_4M$colnames,cex.axis = 0.7, las = 3,range=2,main="Cook's distance ~ SCTLD_RNA_Region_season")
beep()
```




#### Dispersion Plot
```{r}

# ~ SCTLD_RNA_Region_season

# png("results/5_All_4M_DEG/Dispersionplot_dds_sctld.reg.season.png", width=15, height=6, units = "in", res = 300)
plotDispEsts(dds_sctld.reg.season, main="Dispersion plot dds_sctld.reg.season ~ SCTLD_RNA_Region_season")
# dev.off()


beep()
```

#### PCA DESeq2 plot with vst transformation dds_sctld.reg.season
```{r}
#PCA dds_sctld.reg.season:~ SCTLD_RNA_Region_season

pcaData = plotPCA(vst_dds_sctld.reg.season, intgroup=c("colnames",'SCTLD_RNA',"Region","sample_month","time_point","Location", "sample_day","percent_uniq", "season"), 
returnData=TRUE)
percentVar = round(100 * attr(pcaData, "percentVar"))

# png("results/5_All_4M_DEG/PCA_dds_sctld.reg.season_vst.png", width=10, height=8, units = "in", res = 300)
ggplot(pcaData, aes(PC1, PC2, colour = SCTLD_RNA, shape =season )) + 
geom_point(size = 2) + theme_bw() + 
geom_text_repel(aes(label = Region), nudge_x = -1, nudge_y = 0.2, size = 3) +
ggtitle("Principal Component Analysis vst", subtitle = "~ SCTLD_RNA_Region_season") +
xlab(paste0("PC1: ",percentVar[1],"% variance")) +
ylab(paste0("PC2: ",percentVar[2],"% variance"))
# dev.off()

beep()
```

#### PCA DESeq2 plot with rlog transformation
```{r}
#PCA dds:~ ~ SCTLD_RNA_Region_season

pcaData_rlog = plotPCA(rlog_dds_sctld.reg.season, intgroup=c("colnames",'SCTLD_RNA',"Region","sample_month","time_point","Location"), 
returnData=TRUE)
percentVar_rlog = round(100 * attr(pcaData_rlog, "percentVar"))

# png("results/5_All_4M_DEG/PCA_dds_sctld.reg.season_rlog.png", width=10, height=8, units = "in", res = 300)
ggplot(pcaData_rlog, aes(PC1, PC2, colour = Region, shape = SCTLD_RNA)) +
geom_point(size = 2) + theme_bw() +
geom_text_repel(aes(label = sample_month), nudge_x = -1, nudge_y = 0.2, size = 3) +
ggtitle("Principal Component Analysis rlog ", subtitle = "~ SCTLD_RNA_Region_season") +
xlab(paste0("PC1: ",percentVar_rlog[1],"% variance")) +
ylab(paste0("PC2: ",percentVar_rlog[2],"% variance"))

# dev.off()

```

#### Checking sumwintery results for ~ SCTLD_RNA_Region_season to see if there is any concerns about outliers
```{r}

#Results dds_sctld.reg.season:~ SCTLD_RNA_Region_season
res_sctld.reg.season<- results(dds_sctld.reg.season)


```


## Getting list of differencialy expressed genes (DEG) per relevant contrast

- Factor names dds_sctld.reg.season:~ SCTLD_RNA_Region_season
```{r}
resultsNames(dds_sctld.reg.season)
```

### Susceptability effect of gene expression per season and region

#### Contrast for Lower Keys ~ SCTLD_RNA_Region_season
##### DEG btw SCTLD_affected vs SCTLD_ unaffected in fall --> Fall
```{r}
res_lower_affected_fall_vs_unaffected_fall<- results(dds_sctld.reg.season,contrast =
                                        
                              c("SCTLD_RNA_Region_season",'SCTLD_affected_Lower_fall',"SCTLD_unaffected_Lower_fall"))

resOrdered_lower_affected_fall_vs_unaffected_fall<- as.data.frame(
               
                                 res_lower_affected_fall_vs_unaffected_fall[order(res_lower_affected_fall_vs_unaffected_fall$padj),])
                
DEG_lower_fall_aff_vs_unaff <-resOrdered_lower_affected_fall_vs_unaffected_fall %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_lower_affected_fall_vs_unaffected_fall)) %>%
   
                                filter(padj<0.1)


```

##### DEG btw SCTLD_affected vs SCTLD_ unaffected in winter --> winter
```{r}
res_lower_affected_winter_vs_unaffected_winter<- results(dds_sctld.reg.season,contrast =
                                        
                              c("SCTLD_RNA_Region_season",'SCTLD_affected_Lower_winter',"SCTLD_unaffected_Lower_winter"))

resOrdered_lower_affected_winter_vs_unaffected_winter<- as.data.frame(
               
                                 res_lower_affected_winter_vs_unaffected_winter[order(res_lower_affected_winter_vs_unaffected_winter$padj),])
                
DEG_lower_winter_aff_vs_unaff <-resOrdered_lower_affected_winter_vs_unaffected_winter %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_lower_affected_winter_vs_unaffected_winter)) %>%
   
                                filter(padj<0.1)


```


#### Contrast for ECA ~ SCTLD_RNA_Region_season

##### DEG btw SCTLD_affected vs SCTLD_ unaffected in fall --> fall
```{r}
res_ECA_affected_fall_vs_unaffected_fall<- results(dds_sctld.reg.season,contrast =
                                        
                              c("SCTLD_RNA_Region_season",'SCTLD_affected_ECA_fall',"SCTLD_unaffected_ECA_fall"))

resOrdered_ECA_affected_fall_vs_unaffected_fall<- as.data.frame(
               
                                 res_ECA_affected_fall_vs_unaffected_fall[order(res_ECA_affected_fall_vs_unaffected_fall$padj),])
                
DEG_ECA_fall_aff_vs_unaff <-resOrdered_ECA_affected_fall_vs_unaffected_fall %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_ECA_affected_fall_vs_unaffected_fall)) %>%
   
                                filter(padj<0.1)


```

##### DEG btw SCTLD_affected vs SCTLD_ unaffected in winter --> winter
```{r}
res_ECA_affected_winter_vs_unaffected_winter<- results(dds_sctld.reg.season,contrast =
                                        
                              c("SCTLD_RNA_Region_season",'SCTLD_affected_ECA_winter',"SCTLD_unaffected_ECA_winter"))

resOrdered_ECA_affected_winter_vs_unaffected_winter<- as.data.frame(
               
                                 res_ECA_affected_winter_vs_unaffected_winter[order(res_ECA_affected_winter_vs_unaffected_winter$padj),])
                
DEG_ECA_winter_aff_vs_unaff <-resOrdered_ECA_affected_winter_vs_unaffected_winter %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_ECA_affected_winter_vs_unaffected_winter)) %>%
   
                                filter(padj<0.1)
```

#### Creating a data frame for all genes with the logFoldChanges and padj per contrast
```{r}
# ECA affected vs unaffected in fall

ECA_fall_log.padj<- resOrdered_ECA_affected_fall_vs_unaffected_fall %>%
  
                  mutate(ofav_gene = rownames(resOrdered_ECA_affected_fall_vs_unaffected_fall)) %>% 
  
                  mutate(log_ECA_fall = log2FoldChange) %>% 
  
                  mutate(padj_ECA_fall = padj) %>% 
  
                  mutate(baseMean_ECA_fall = baseMean) %>%
  
                  select(ofav_gene, log_ECA_fall, padj_ECA_fall,baseMean_ECA_fall)  

# ECA affected vs unaffected in winter
ECA_winter_log.padj<- resOrdered_ECA_affected_winter_vs_unaffected_winter %>%
  
                  mutate(ofav_gene = rownames(resOrdered_ECA_affected_winter_vs_unaffected_winter)) %>% 
  
                  mutate(log_ECA_winter = log2FoldChange) %>% 
  
                  mutate(padj_ECA_winter = padj) %>% 
 
                  mutate(baseMean_ECA_winter = baseMean) %>%
  
                  select(ofav_gene, log_ECA_winter, padj_ECA_winter,baseMean_ECA_winter)
                  
# Lower affected vs unaffected in fall

lower_fall_log.padj<- resOrdered_lower_affected_fall_vs_unaffected_fall %>%
  
                  mutate(ofav_gene = rownames(resOrdered_lower_affected_fall_vs_unaffected_fall)) %>% 
  
                  mutate(log_lower_fall = log2FoldChange) %>% 
  
                  mutate(padj_lower_fall = padj) %>% 
              
                  mutate(baseMean_lower_fall = baseMean) %>%
  
                  select(ofav_gene, log_lower_fall, padj_lower_fall,baseMean_lower_fall)   
                  
lower_winter_log.padj<- resOrdered_lower_affected_winter_vs_unaffected_winter %>%
  
                  mutate(ofav_gene = rownames(resOrdered_lower_affected_winter_vs_unaffected_winter)) %>% 
  
                  mutate(log_lower_winter = log2FoldChange) %>% 
  
                  mutate(padj_lower_winter = padj) %>% 
  
                  mutate(baseMean_lower_winter = baseMean) %>%
  
                  select(ofav_gene, log_lower_winter, padj_lower_winter,baseMean_lower_winter)
```

#### Creating data frame with all the results and all the genes
```{r}
ECA.affvsunaff_season<- ECA_fall_log.padj %>% inner_join(ECA_winter_log.padj,by="ofav_gene")

lower.affvsunaff_season<- lower_fall_log.padj %>% inner_join(lower_winter_log.padj,by="ofav_gene")

suscept_season_region<- ECA.affvsunaff_season %>% inner_join(lower.affvsunaff_season,by="ofav_gene")


# DEG in fall in ECA
suscept_season_region$sig_ECA_fall <- ifelse(suscept_season_region$padj_ECA_fall < 0.1, "TRUE", "FALSE") #--> 2

# DEG in winter in ECA
suscept_season_region$sig_ECA_winter <- ifelse(suscept_season_region$padj_ECA_winter < 0.1, "TRUE", "FALSE") #--> 89

# DEG in winter in Lower
suscept_season_region$sig_lower_winter <- ifelse(suscept_season_region$padj_lower_winter < 0.1, "TRUE", "FALSE") #--> 32

# DEG in fall in Lower
suscept_season_region$sig_lower_fall <- ifelse(suscept_season_region$padj_lower_fall < 0.1, "TRUE", "FALSE") #--> 0


# DEG in fall in both regions
suscept_season_region$sig_region_fall <- ifelse(suscept_season_region$padj_ECA_fall < 0.1 & suscept_season_region$padj_lower_fall < 0.1, "TRUE", "FALSE") #--> 0

# DEG in winter in both regions
suscept_season_region$sig_region_winter <- ifelse(suscept_season_region$padj_ECA_winter < 0.1 & suscept_season_region$padj_lower_winter < 0.1, "TRUE", "FALSE") #--> 1

# DEG in ECA in both seasons --> 1
suscept_season_region$sig_ECA_2seasons <- ifelse(suscept_season_region$padj_ECA_winter < 0.1 & suscept_season_region$padj_ECA_fall < 0.1, "TRUE", "FALSE") # --> 1

# DEG in Lower in both seasons --> 1
suscept_season_region$sig_lower_2seasons <- ifelse(suscept_season_region$padj_lower_winter < 0.1 & suscept_season_region$padj_lower_fall < 0.1, "TRUE", "FALSE") # --> 0

# DEG in at least in one contrast
suscept_season_region$sig_genes.affvsunaff_region_season <- ifelse(
  
  suscept_season_region$padj_ECA_fall < 0.1 |
  
  suscept_season_region$padj_ECA_winter < 0.1 |
  
  suscept_season_region$padj_lower_fall < 0.1 |
  
  suscept_season_region$padj_lower_winter < 0.1,
  
  "TRUE", "FALSE") # --> 51

suscept_season_region$sig_genes.affvsunaff_region_winter<- 
  
  ifelse(suscept_season_region$padj_ECA_winter < 0.1 &
                                                              suscept_season_region$padj_lower_winter < 0.1, "TRUE", "FALSE")

suscept_season_region$sig_genes.affvsunaff_region_winter<- as.logical(suscept_season_region$sig_genes.affvsunaff_region_winter) 
  
sum(suscept_season_region$sig_genes.affvsunaff_region_winter)#-->0

suscept_season_region$sig_genes.affvsunaff_region_season<- as.logical(suscept_season_region$sig_genes.affvsunaff_region_season)

total_sig_season_region <- sum(suscept_season_region$sig_genes.affvsunaff_region_season) # -->51

```

#### Creating data frame with all the results and all the genes : UP or down regulater
```{r}

suscept_season_region$DEG_ECA_winter<- "NO"
suscept_season_region$DEG_ECA_winter[suscept_season_region$log_ECA_winter > 0 & suscept_season_region$padj_ECA_winter < 0.1] <- "UP"
suscept_season_region$DEG_ECA_winter[suscept_season_region$log_ECA_winter < 0 & suscept_season_region$padj_ECA_winter < 0.1] <- "DOWN"

suscept_season_region$DEG_ECA_fall<- "NO"
suscept_season_region$DEG_ECA_fall[suscept_season_region$log_ECA_fall > 0 & suscept_season_region$padj_ECA_fall < 0.1] <- "UP"
suscept_season_region$DEG_ECA_fall[suscept_season_region$log_ECA_fall < 0 & suscept_season_region$padj_ECA_fall < 0.1] <- "DOWN"

suscept_season_region$DEG_lower_winter<- "NO"
suscept_season_region$DEG_lower_winter[suscept_season_region$log_lower_winter > 0 & suscept_season_region$padj_lower_winter < 0.1] <- "UP"
suscept_season_region$DEG_lower_winter[suscept_season_region$log_lower_winter < 0 & suscept_season_region$padj_lower_winter < 0.1] <- "DOWN"

suscept_season_region$DEG_lower_fall<- "NO"
suscept_season_region$DEG_lower_fall[suscept_season_region$log_lower_fall > 0 & suscept_season_region$padj_lower_fall < 0.1] <- "UP"
suscept_season_region$DEG_lower_fall[suscept_season_region$log_lower_fall < 0 & suscept_season_region$padj_lower_fall < 0.1] <- "DOWN"


```

```{r}
suscept_season_region_anno<- suscept_season_region %>% left_join(anno, by="ofav_gene")
```


#### Extracting DEG
```{r}
sig_suscept_season_region<- suscept_season_region_anno %>% filter(sig_genes.affvsunaff_region_season == TRUE)#-->191

sig_suscept_season_region_w<- sig_suscept_season_region %>% select(-gDNA,-mRNA, -"CDS-transcript",-Translation)

sig_genes.affvsunaff_winter_ECA<- sig_suscept_season_region_w %>% filter(sig_ECA_winter == TRUE) #-->160

sig_genes.affvsunaff_fall_lower<- sig_suscept_season_region_w %>% filter(sig_lower_fall == TRUE) #-->0

sig_genes.affvsunaff_winter_lower<- sig_suscept_season_region_w %>% filter(sig_lower_winter == TRUE)#-->31

# sig_genes.affvsunaff_fall_ECA--> 0 DEG
```

#### Counts per contrast and all sig genes
```{r}
# VST counts for winter ECA DEG

ids_winter_ECA<- sig_genes.affvsunaff_winter_ECA$ofav_gene

count_winter_ECA_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_winter_ECA),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_winter_ECA_vst))
all(rownames(colData_4M) == colnames(count_winter_ECA_vst))

count_winter_ECA_vst_df<-as.data.frame( assay(count_winter_ECA_vst))

# VST counts for winter lower DEG --> 14

ids_winter_lower<- sig_genes.affvsunaff_winter_lower$ofav_gene

count_winter_lower_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_winter_lower),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_winter_lower_vst))
all(rownames(colData_4M) == colnames(count_winter_lower_vst))

count_winter_lower_vst_df<-as.data.frame( assay(count_winter_lower_vst))

# VST counts for all DEG affected vs unaffected

ids_region_2seasons<- sig_suscept_season_region$ofav_gene

count_region_2seasons_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_region_2seasons),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_region_2seasons_vst))
all(rownames(colData_4M) == colnames(count_region_2seasons_vst))

count_winter_lower_vst_df<-as.data.frame( assay(count_region_2seasons_vst))

# VST counts for fall ECA DEG --> only 1 DEG

# VST counts for fall lower DEG --> NO DEGs
```

### Writing tables and data counts
#### vst transformed read counts with different DEG lists
```{r}
# write_rds(count_region_2seasons_vst, "cache/count_region_2seasons_vst.rds")

```

#### Writing results tables
```{r}

# write_csv(sig_suscept_season_region_w, "results/5_All_4M_DEG/sig_suscept_season_region.csv")
# 
# write_csv(sig_genes.WvsF_winter_ECA, "results/5_All_4M_DEG/sig_genes.WvsF_winter_ECA.csv")
# 
# write_csv(sig_genes.WvsF_fall_ECA, "results/5_All_4M_DEG/sig_genes.WvsF_fall_ECA.csv")
# 
# write_csv(sig_genes.WvsF_winter_lower, "results/5_All_4M_DEG/sig_genes.WvsF_winter_lower.csv")

```


#### Plots susceptibility effect in gene expression per season and region

##### PCA plots
###### PCA sign genes ECA winter 
```{r}
pcaData = plotPCA(count_winter_ECA_vst, intgroup=c("colnames",'SCTLD_RNA',"Region","sample_month","time_point","Location","SCTLD_RNA_region","SCTLD_RNA_SP_region", "sample_day","percent_uniq","Sample_ID","SCTLD_RNA_SP","SP_region","Core_Site_Location","Histopath_supported_resistance","SCTLD_RNA3","sample_period","Plate_Number","Disease_on_colony_when_coring","SCTLD_RNA_Region_season"), 
returnData=TRUE)

pcaData$SCTLD_RNA<- as.factor(pcaData$SCTLD_RNA)
pcaData$SCTLD_RNA<- factor(pcaData$SCTLD_RNA,levels=c("SCTLD_affected","SCTLD_unaffected"))

pcaData$SP_region<- as.factor(pcaData$SP_region)
pcaData$SP_region<- factor(pcaData$SP_region,levels=c("SP2_ECA","SP2_Lower","SP3_ECA","SP3_Lower"))
                                

pcaData$SCTLD_RNA_region<- as.factor(pcaData$SCTLD_RNA_region)
pcaData$SCTLD_RNA_region<- factor(pcaData$SCTLD_RNA_region,levels=c("SCTLD_affected_ECA","SCTLD_unaffected_ECA","SCTLD_affected_Lower","SCTLD_unaffected_Lower"))

pcaData$SCTLD_RNA_SP_region<- as.factor(pcaData$SCTLD_RNA_SP_region)
pcaData$SCTLD_RNA_SP_region<- factor(pcaData$SCTLD_RNA_SP_region,levels=c("SCTLD_affected_SP3_ECA","SCTLD_unaffected_SP3_ECA","SCTLD_affected_SP3_Lower","SCTLD_unaffected_SP3_Lower","SCTLD_affected_SP2_ECA","SCTLD_unaffected_SP2_ECA","SCTLD_affected_SP2_Lower","SCTLD_unaffected_SP2_Lower"))
pcaData$SCTLD_RNA_Region_season<- as.factor(pcaData$SCTLD_RNA_Region_season)
pcaData$SCTLD_RNA_Region_season<- factor(pcaData$SCTLD_RNA_Region_season,levels=c("SCTLD_affected_Lower_fall","SCTLD_unaffected_Lower_fall","SCTLD_affected_ECA_fall","SCTLD_unaffected_ECA_fall","SCTLD_affected_Lower_winter","SCTLD_unaffected_Lower_winter","SCTLD_affected_ECA_winter","SCTLD_unaffected_ECA_winter"))

percentVar = round(100 * attr(pcaData, "percentVar"))


class_colors <- c("SCTLD_affected_ECA_winter" ="#F2994A" ,  # Grey
                   "SCTLD_unaffected_ECA_winter" = "#2D9CDB",  # Light Blue
                   "SCTLD_affected_Lower_fall" = "#D3D3D3",  # Teal
                   "SCTLD_unaffected_Lower_fall" = "#D3D3D3",
                   "SCTLD_affected_ECA_fall" = "#D3D3D3",  # Teal
                   "SCTLD_unaffected_ECA_fall" = "#D3D3D3",
                   "SCTLD_affected_Lower_winter" = "#D3D3D3",  # Teal
                   "SCTLD_unaffected_Lower_winter" = "#D3D3D3")  # Orange

# class_colors <- c("1" ="#009E73",  # Teal
#                    "0" = "#E69F00")

png("results/DEG/PCA_ECA_winter_affvsunaff.png", width=10, height=10, units = "in", res = 300)
ggplot(pcaData, aes(PC1, PC2, colour = SCTLD_RNA_Region_season, shape = SCTLD_RNA)) + 
geom_point(size = 3) + 
stat_ellipse(aes(group = SCTLD_RNA_Region_season),type = "t", linetype = "dashed", level = 0.95) +
geom_text_repel(aes(label = Sample_ID), nudge_x = -1, nudge_y = 0.2, size = 3, max.overlaps = 100) +
scale_color_manual(values = class_colors) +
ggtitle("PCA DESeq2:vst  DEG between aff and unaff colonies in ECA region in winter") +
xlab(paste0("PC1: ",percentVar[1],"% variance")) +
ylab(paste0("PC2: ",percentVar[2],"% variance")) +
theme_bw() + 
theme_minimal()
# theme(panel.background = element_rect(fill = "lightgrey"),legend.position = "top")
dev.off()

beep()
```
###### PCA sign genes Lower winter 
```{r}
pcaData = plotPCA(count_winter_lower_vst, intgroup=c("colnames",'SCTLD_RNA',"Region","sample_month","time_point","Location","SCTLD_RNA_region","SCTLD_RNA_SP_region", "sample_day","percent_uniq","Sample_ID","SCTLD_RNA_SP","SP_region","Core_Site_Location","Histopath_supported_resistance","SCTLD_RNA3","sample_period","Plate_Number","Disease_on_colony_when_coring","SCTLD_RNA_Region_season"), 
returnData=TRUE)

pcaData$SCTLD_RNA<- as.factor(pcaData$SCTLD_RNA)
pcaData$SCTLD_RNA<- factor(pcaData$SCTLD_RNA,levels=c("SCTLD_affected","SCTLD_unaffected"))

pcaData$SP_region<- as.factor(pcaData$SP_region)
pcaData$SP_region<- factor(pcaData$SP_region,levels=c("SP2_ECA","SP2_Lower","SP3_ECA","SP3_Lower"))
                                

pcaData$SCTLD_RNA_region<- as.factor(pcaData$SCTLD_RNA_region)
pcaData$SCTLD_RNA_region<- factor(pcaData$SCTLD_RNA_region,levels=c("SCTLD_affected_ECA","SCTLD_unaffected_ECA","SCTLD_affected_Lower","SCTLD_unaffected_Lower"))

pcaData$SCTLD_RNA_SP_region<- as.factor(pcaData$SCTLD_RNA_SP_region)
pcaData$SCTLD_RNA_SP_region<- factor(pcaData$SCTLD_RNA_SP_region,levels=c("SCTLD_affected_SP3_ECA","SCTLD_unaffected_SP3_ECA","SCTLD_affected_SP3_Lower","SCTLD_unaffected_SP3_Lower","SCTLD_affected_SP2_ECA","SCTLD_unaffected_SP2_ECA","SCTLD_affected_SP2_Lower","SCTLD_unaffected_SP2_Lower"))
pcaData$SCTLD_RNA_Region_season<- as.factor(pcaData$SCTLD_RNA_Region_season)
pcaData$SCTLD_RNA_Region_season<- factor(pcaData$SCTLD_RNA_Region_season,levels=c("SCTLD_affected_Lower_fall","SCTLD_unaffected_Lower_fall","SCTLD_affected_ECA_fall","SCTLD_unaffected_ECA_fall","SCTLD_affected_Lower_winter","SCTLD_unaffected_Lower_winter","SCTLD_affected_ECA_winter","SCTLD_unaffected_ECA_winter"))

percentVar = round(100 * attr(pcaData, "percentVar"))


class_colors <- c("SCTLD_affected_ECA_winter" ="#D3D3D3" ,  # Grey
                   "SCTLD_unaffected_ECA_winter" = "#D3D3D3",  # Light Blue
                   "SCTLD_affected_Lower_fall" = "#D3D3D3",  # Teal
                   "SCTLD_unaffected_Lower_fall" = "#D3D3D3",
                   "SCTLD_affected_ECA_fall" = "#D3D3D3",  # Teal
                   "SCTLD_unaffected_ECA_fall" = "#D3D3D3",
                   "SCTLD_affected_Lower_winter" = "#F2994A",  # Teal
                   "SCTLD_unaffected_Lower_winter" = "#2D9CDB")  # Orange

png("results/DEG/PCA_lower_winter_affvsunaff.png", width=10, height=10, units = "in", res = 300)
ggplot(pcaData, aes(PC1, PC2, colour = SCTLD_RNA_Region_season, shape = SCTLD_RNA)) + 
geom_point(size = 3) + 
stat_ellipse(aes(group = SCTLD_RNA_Region_season),type = "t", linetype = "dashed", level = 0.95) +
geom_text_repel(aes(label = Sample_ID), nudge_x = -1, nudge_y = 0.2, size = 3) +
scale_color_manual(values = class_colors) +
ggtitle("PCA DESeq2:vst , DEG between aff and unaff colonies in Lower Keys in winter") +
xlab(paste0("PC1: ",percentVar[1],"% variance")) +
ylab(paste0("PC2: ",percentVar[2],"% variance")) +
theme_bw() + 
theme_minimal()
# theme(panel.background = element_rect(fill = "lightgrey"),legend.position = "top")
dev.off()

beep()
```
###### PCA sign genes overall
```{r}
pcaData = plotPCA(count_region_2seasons_vst, intgroup=c("colnames",'SCTLD_RNA',"Region","sample_month","time_point","Location","SCTLD_RNA_region","SCTLD_RNA_SP_region", "sample_day","percent_uniq","Sample_ID","SCTLD_RNA_SP","SP_region","Core_Site_Location","Histopath_supported_resistance","SCTLD_RNA3","sample_period","Plate_Number","Disease_on_colony_when_coring","SCTLD_RNA_Region_season"), 
returnData=TRUE)

pcaData$SCTLD_RNA<- as.factor(pcaData$SCTLD_RNA)
pcaData$SCTLD_RNA<- factor(pcaData$SCTLD_RNA,levels=c("SCTLD_affected","SCTLD_unaffected"))

pcaData$SP_region<- as.factor(pcaData$SP_region)
pcaData$SP_region<- factor(pcaData$SP_region,levels=c("SP2_ECA","SP2_Lower","SP3_ECA","SP3_Lower"))
                                

pcaData$SCTLD_RNA_region<- as.factor(pcaData$SCTLD_RNA_region)
pcaData$SCTLD_RNA_region<- factor(pcaData$SCTLD_RNA_region,levels=c("SCTLD_affected_ECA","SCTLD_unaffected_ECA","SCTLD_affected_Lower","SCTLD_unaffected_Lower"))

pcaData$SCTLD_RNA_SP_region<- as.factor(pcaData$SCTLD_RNA_SP_region)
pcaData$SCTLD_RNA_SP_region<- factor(pcaData$SCTLD_RNA_SP_region,levels=c("SCTLD_affected_SP3_ECA","SCTLD_unaffected_SP3_ECA","SCTLD_affected_SP3_Lower","SCTLD_unaffected_SP3_Lower","SCTLD_affected_SP2_ECA","SCTLD_unaffected_SP2_ECA","SCTLD_affected_SP2_Lower","SCTLD_unaffected_SP2_Lower"))

pcaData$SCTLD_RNA_Region_season<- as.factor(pcaData$SCTLD_RNA_Region_season)
pcaData$SCTLD_RNA_Region_season<- factor(pcaData$SCTLD_RNA_Region_season,levels=c("SCTLD_affected_Lower_fall","SCTLD_unaffected_Lower_fall","SCTLD_affected_ECA_fall","SCTLD_unaffected_ECA_fall","SCTLD_affected_Lower_winter","SCTLD_unaffected_Lower_winter","SCTLD_affected_ECA_winter","SCTLD_unaffected_ECA_winter"))

percentVar = round(100 * attr(pcaData, "percentVar"))


class_colors <- c("SCTLD_affected_ECA_winter" ="#F2994A" ,  # Grey
                   "SCTLD_unaffected_ECA_winter" = "#2D9CDB",  # Light Blue
                   "SCTLD_affected_Lower_fall" = "#D3D3D3",  # Teal
                   "SCTLD_unaffected_Lower_fall" = "#D3D3D3",
                   "SCTLD_affected_ECA_fall" = "#D3D3D3",  # Teal
                   "SCTLD_unaffected_ECA_fall" = "#D3D3D3",
                   "SCTLD_affected_Lower_winter" = "#F2994A",  # Teal
                   "SCTLD_unaffected_Lower_winter" = "#2D9CDB")  

# png("results/8_mint/PCA_mint_comp1.png", width=10, height=10, units = "in", res = 300)
ggplot(pcaData, aes(PC1, PC2, colour = SCTLD_RNA_Region_season, shape = SCTLD_RNA)) + 
geom_point(size = 3) + 
stat_ellipse(aes(group = SCTLD_RNA_Region_season),type = "t", linetype = "dashed", level = 0.95) +
geom_text_repel(aes(label = Sample_ID), nudge_x = -1, nudge_y = 0.2, size = 3, max.overlaps = 100) +
scale_color_manual(values = class_colors) +
ggtitle("PCA DESeq2:vst , Lower and ECA per season DEG ~ SCTLD_RNA_Region_season") +
xlab(paste0("PC1: ",percentVar[1],"% variance")) +
ylab(paste0("PC2: ",percentVar[2],"% variance")) +
theme_bw() + 
theme_minimal()
# theme(panel.background = element_rect(fill = "lightgrey"),legend.position = "top")
# dev.off()

beep()
```


##### Heatmaps for susceptibility effect per season and region

######DEG ECA winter
```{r}
gene_ECA_winter<- sig_genes.affvsunaff_winter_ECA
rownames(gene_ECA_winter)<- gene_ECA_winter$ofav_gene
# gene_ECA_winter<- gene_ECA_winter[order(gene_ECA_winter$log_ECA_winter, decreasing = TRUE),]

ECA_winter<- colData_4M %>% filter(SP_region == "SP3_ECA")
samples_ECA_winter<- ECA_winter$Sample_ID


vst_out_ECA_winter<-count_winter_ECA_vst[(rownames(count_winter_ECA_vst) %in% gene_ECA_winter),]

vst_out_ECA_winter<-count_winter_ECA_vst[,(colnames(count_winter_ECA_vst) %in% samples_ECA_winter)]

# making the rownames and column names identical
all(rownames(samples_ECA_winter) %in% colnames(vst_out_ECA_winter))
all(rownames(samples_ECA_winter) == colnames(vst_out_ECA_winter))

mat_ECA_winter<-assay(vst_out_ECA_winter) #sig genes x samples
mat_ECA_winter.scaled <- t(apply(mat_ECA_winter, 1, scale)) #center and scale each column (Z-score) then transpose
colnames(mat_ECA_winter.scaled)<-colnames(mat_ECA_winter)

suscept_ECA_winter <- as.factor(ECA_winter$SCTLD_RNA)
suscept_ECA_winter <- factor(suscept_ECA_winter,levels = c("SCTLD_affected", "SCTLD_unaffected"))

region_ECA_winter <- as.factor(ECA_winter$Region)
region_ECA_winter <- factor(region_ECA_winter,levels = c("ECA"))

SP_ECA_winter <- as.factor(ECA_winter$sample_period)
SP_ECA_winter <- factor(SP_ECA_winter,levels = c("SP3"))


ha_ECA_winter <- HeatmapAnnotation( 
                         Winter  = SP_ECA_winter,
                         ECA = region_ECA_winter,
                         SCTLD = suscept_ECA_winter,
                         col = list(
                           Winter = c("SP3"="#0072B2"),
                           ECA = c("ECA"="#009E73"),
                           SCTLD = c("SCTLD_affected" = "#E69F00", "SCTLD_unaffected"="#2D9CDB")
                           ),
                         annotation_legend_param = list(
                           Winter = NA,
                           ECA = NA,
                           SCTLD = list(title = "SCTLD susceptibility"),
                          
                           show_legend = c(Winter = FALSE, ECA = FALSE,Winter = TRUE)) 
                         )
  # sample_data = anno_barplot(mat_ECA_winter, gp = gpar(fill = "blue"), height = unit(2, "cm"))  # Barplot for samples


```

```{r}
png("results/DEG/heatmap_ECA_winter_affvsunaff.png", width=15, height=10, units = "in", res = 300)
Heatmap(mat_ECA_winter.scaled, 
        cluster_columns =  T,  # Disable clustering for columns
        cluster_rows =T, 
        show_row_names = FALSE,
        top_annotation = ha_ECA_winter)
dev.off()
```



######DEG lower winter
```{r}
gene_lower_winter<- sig_genes.affvsunaff_winter_lower
rownames(gene_lower_winter)<- gene_lower_winter$ofav_gene
# gene_lower_winter<- gene_lower_winter[order(gene_lower_winter$log_lower_winter, decreasing = TRUE),]

lower_winter<- colData_4M %>% filter(SP_region == "SP3_Lower")
samples_lower_winter<- lower_winter$Sample_ID


vst_out_lower_winter<-count_winter_lower_vst[(rownames(count_winter_lower_vst) %in% gene_lower_winter),]

vst_out_lower_winter<-count_winter_lower_vst[,(colnames(count_winter_lower_vst) %in% samples_lower_winter)]

# making the rownames and column names identical
all(rownames(samples_lower_winter) %in% colnames(vst_out_lower_winter))
all(rownames(samples_lower_winter) == colnames(vst_out_lower_winter))

mat_lower_winter<-assay(vst_out_lower_winter) #sig genes x samples
mat_lower_winter.scaled <- t(apply(mat_lower_winter, 1, scale)) #center and scale each column (Z-score) then transpose
colnames(mat_lower_winter.scaled)<-colnames(mat_lower_winter)

suscept_lower_winter <- as.factor(lower_winter$SCTLD_RNA)
suscept_lower_winter <- factor(suscept_lower_winter,levels = c("SCTLD_affected", "SCTLD_unaffected"))

region_lower_winter <- as.factor(lower_winter$Region)
region_lower_winter <- factor(region_lower_winter,levels = c("Lower"))

SP_lower_winter <- as.factor(lower_winter$sample_period)
SP_lower_winter <- factor(SP_lower_winter,levels = c("SP3"))


ha_lower_winter <- HeatmapAnnotation( 
                         Winter  = SP_lower_winter,
                         Lower = region_lower_winter,
                         SCTLD = suscept_lower_winter,
                         col = list(
                           Winter = c("SP3"="#0072B2"),
                           lower = c("Lower"="#F0E442"),
                           SCTLD = c("SCTLD_affected" = "#E69F00", "SCTLD_unaffected"="#2D9CDB")
                           ),
                         annotation_legend_param = list(
                           Winter = NA,
                           Lower = NA,
                           SCTLD = list(title = "SCTLD susceptibility")),
                           show_legend = c(Winter = FALSE, lower = FALSE, SCTLD = TRUE) 
                         )
  # sample_data = anno_barplot(mat_lower_winter, gp = gpar(fill = "blue"), height = unit(2, "cm"))  # Barplot for samples


```

```{r}
png("results/DEG/heatmap_lower_winter_affvsunaff.png", width=15, height=10, units = "in", res = 300)

Heatmap(mat_lower_winter.scaled, 
        cluster_columns =  T,  # Disable clustering for columns
        cluster_rows =T,     
        top_annotation = ha_lower_winter)
dev.off()
```


###### All DEG Season Region
```{r}
gene_winter_region<- sig_suscept_season_region
rownames(gene_winter_region)<- gene_winter_region$ofav_gene
# gene_winter_region<- gene_winter_region[order(gene_winter_region$log_winter_region, decreasing = TRUE),]

winter_region<- colData_4M %>% filter(sample_period == "SP3")
samples_winter_region<- winter_region$Sample_ID


vst_out_winter_region<-count_winter_lower_vst[(rownames(count_winter_lower_vst) %in% gene_winter_region),]

vst_out_winter_region<-count_winter_lower_vst[,(colnames(count_winter_lower_vst) %in% samples_winter_region)]

# making the rownames and column names identical
all(rownames(samples_winter_region) %in% colnames(vst_out_winter_region))
all(rownames(samples_winter_region) == colnames(vst_out_winter_region))

mat_winter_region<-assay(vst_out_winter_region) #sig genes x samples
mat_winter_region.scaled <- t(apply(mat_winter_region, 1, scale)) #center and scale each column (Z-score) then transpose
colnames(mat_winter_region.scaled)<-colnames(mat_winter_region)

suscept_winter_region <- as.factor(winter_region$SCTLD_RNA)
suscept_winter_region <- factor(suscept_winter_region,levels = c("SCTLD_affected", "SCTLD_unaffected"))

region_winter_region <- as.factor(winter_region$Region)
region_winter_region <- factor(region_winter_region,levels = c("ECA","Lower"))

SP_winter_region <- as.factor(winter_region$sample_period)
SP_winter_region <- factor(SP_winter_region,levels = c("SP3"))


ha_winter_region <- HeatmapAnnotation( 
                         Winter  = SP_winter_region,
                         Rgion = region_winter_region,
                         SCTLD = suscept_winter_region,
                         col = list(
                           Winter = c("SP3"="#0072B2"),
                           Rgion = c("ECA"="#009E73","Lower"="#F0E442"),
                           SCTLD = c("SCTLD_affected" = "#E69F00", "SCTLD_unaffected"="#2D9CDB")
                           ),
                         annotation_legend_param = list(
                           Winter = NA,
                           Rgion = list(title = "Region"),
                           SCTLD = list(title = "SCTLD susceptibility")),
                           show_legend = c(Winter = FALSE, Rgion = TRUE, SCTLD = TRUE) 
                         )
  # sample_data = anno_barplot(mat_winter_region, gp = gpar(fill = "blue"), height = unit(2, "cm"))  # Barplot for samples

```

```{r}
Heatmap(mat_winter_region.scaled, 
        cluster_columns =  T,  # Disable clustering for columns
        cluster_rows =T,     
        top_annotation = ha_winter_region)
```
###### Printing out heatmaps
```{r}
# png("results/heatmap_v1.png", res = 300, width = 5000, height = 5500)
# print(h)
# dev.off()
```










































































## Season effect of gene expression per susceptibility and region:Gene expression Winter vs Fall

- we will use wintersamples as the "treatment" and fall as the "control"

#### Contrast for Lower Keys ~ SCTLD_RNA_Region_season
```{r}
resultsNames(dds_sctld.reg.season)

```

##### DEG btw SCTLD_affected in fall vs SCTLD_affected in winter--> DEG Lower affected Winter vs Fall

```{r}
res_lower_affected_WvsF<- results(dds_sctld.reg.season,contrast =
                                        
                              c("SCTLD_RNA_Region_season",'SCTLD_affected_Lower_winter',"SCTLD_affected_Lower_fall"))

resOrdered_lower_affected_WvsF<- as.data.frame(
               
                                 res_lower_affected_WvsF[order(res_lower_affected_WvsF$padj),])
                
DEG_lower_aff_WvsF <-resOrdered_lower_affected_WvsF %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_lower_affected_WvsF)) %>%
   
                                filter(padj<0.1)


```

##### DEG btw SCTLD_unaffected in fall vs SCTLD_unaffected in winter--> DEG Lower unaffected Winter vs Fall
```{r}
res_lower_unaffected_WvsF<- results(dds_sctld.reg.season,contrast =
                                        
                              c("SCTLD_RNA_Region_season",'SCTLD_unaffected_Lower_winter',"SCTLD_unaffected_Lower_fall"))

resOrdered_lower_unaffected_WvsF<- as.data.frame(
               
                                 res_lower_unaffected_WvsF[order(res_lower_unaffected_WvsF$padj),])
                
DEG_lower_unaff_WvsF <-resOrdered_lower_unaffected_WvsF %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_lower_unaffected_WvsF)) %>%
   
                                filter(padj<0.1)

```


#### Contrast ECA ~ SCTLD_RNA_Region_season

##### DEG btw SCTLD_affected in fall vs SCTLD_affected in winter--> DEG ECA affected Winter vs Fall
```{r}
res_ECA_affected_WvsF<- results(dds_sctld.reg.season,contrast =
                                        
                              c("SCTLD_RNA_Region_season",'SCTLD_affected_ECA_winter',"SCTLD_affected_ECA_fall"))

resOrdered_ECA_affected_WvsF<- as.data.frame(
               
                                 res_ECA_affected_WvsF[order(res_ECA_affected_WvsF$padj),])
                
DEG_ECA_aff_WvsF <-resOrdered_ECA_affected_WvsF %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_ECA_affected_WvsF)) %>%
   
                                filter(padj<0.1)

```



##### DEG btw SCTLD_unaffected in fall vs SCTLD_unaffected in winter--> DEG ECA unaffected Winter vs Fall
```{r}
res_ECA_unaffected_WvsF<- results(dds_sctld.reg.season,contrast =
                                        
                              c("SCTLD_RNA_Region_season",'SCTLD_unaffected_ECA_winter',"SCTLD_unaffected_ECA_fall"))

resOrdered_ECA_unaffected_WvsF<- as.data.frame(
               
                                 res_ECA_unaffected_WvsF[order(res_ECA_unaffected_WvsF$padj),])
                
DEG_ECA_unaff_WvsF <- resOrdered_ECA_unaffected_WvsF %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_ECA_unaffected_WvsF)) %>%
   
                                filter(padj<0.1)
```



#### Creating a data frame with the logFoldChanges and padj per contrast
```{r}
# ECA affected and unaffected

ECA_aff_WvsF_log.padj<- resOrdered_ECA_affected_WvsF %>%
  
                  mutate(ofav_gene = rownames(resOrdered_ECA_affected_WvsF)) %>% 
  
                  mutate(log_aff_ECA = log2FoldChange) %>% 
  
                  mutate(padj_aff_ECA = padj) %>% 
  
                  mutate(baseMean_aff_ECA = baseMean) %>%
  
                  select(ofav_gene, log_aff_ECA, padj_aff_ECA,baseMean_aff_ECA)   
                  
ECA_unaff_WvsF_log.padj<- resOrdered_ECA_unaffected_WvsF %>%
  
                  mutate(ofav_gene = rownames(resOrdered_ECA_unaffected_WvsF)) %>% 
  
                  mutate(log_unaff_ECA = log2FoldChange) %>% 
  
                  mutate(padj_unaff_ECA = padj) %>% 
 
                  mutate(baseMean_unaff_ECA = baseMean) %>%
  
                  select(ofav_gene, log_unaff_ECA, padj_unaff_ECA,baseMean_unaff_ECA)
                  
# Lower affected and unaffected

lower_aff_WvsF_log.padj<- resOrdered_lower_affected_WvsF %>%
  
                  mutate(ofav_gene = rownames(resOrdered_lower_affected_WvsF)) %>% 
  
                  mutate(log_aff_lower = log2FoldChange) %>% 
  
                  mutate(padj_aff_lower = padj) %>% 
              
                  mutate(baseMean_aff_lower = baseMean) %>%
  
                  select(ofav_gene, log_aff_lower, padj_aff_lower,baseMean_aff_lower)   
                  
lower_unaff_WvsF_log.padj<- resOrdered_lower_unaffected_WvsF %>%
  
                  mutate(ofav_gene = rownames(resOrdered_lower_unaffected_WvsF)) %>% 
  
                  mutate(log_unaff_lower = log2FoldChange) %>% 
  
                  mutate(padj_unaff_lower = padj) %>% 
  
                  mutate(baseMean_unaff_lower = baseMean) %>%
  
                  select(ofav_gene, log_unaff_lower, padj_unaff_lower,baseMean_unaff_lower)

# Creating data fram with all the results
ECA_WvsF<- ECA_aff_WvsF_log.padj %>% inner_join(ECA_unaff_WvsF_log.padj,by="ofav_gene")
lower_WvsF<- lower_aff_WvsF_log.padj %>% inner_join(lower_unaff_WvsF_log.padj,by="ofav_gene")
```


```{r}
region_WvsF<- ECA_WvsF %>% inner_join(lower_WvsF,by="ofav_gene")
```


```{r}
# All DEG WvsF
region_WvsF$sig_genes.WvsF<- "FALSE"

region_WvsF$sig_genes.WvsF[region_WvsF$padj_unaff_ECA< 0.1 ]<- "TRUE"
region_WvsF$sig_genes.WvsF[region_WvsF$padj_aff_ECA< 0.1 ]<- "TRUE"
region_WvsF$sig_genes.WvsF[region_WvsF$padj_unaff_lower< 0.1 ]<- "TRUE"
region_WvsF$sig_genes.WvsF[region_WvsF$padj_aff_lower< 0.1 ]<- "TRUE"


```

```{r}
# DEG unaff region WvsF --> 2576
region_WvsF$sig_unaff.region_WvsF<- "FALSE"

region_WvsF$sig_unaff.region_WvsF[region_WvsF$padj_unaff_ECA< 0.1 ]<- "TRUE"
region_WvsF$sig_unaff.region_WvsF[region_WvsF$padj_unaff_lower< 0.1 ]<- "TRUE"


# DEG aff region WvsF -->3779
region_WvsF$sig_aff.region_WvsF<- "FALSE"

region_WvsF$sig_aff.region_WvsF[region_WvsF$padj_aff_ECA< 0.1 ]<- "TRUE"
region_WvsF$sig_aff.region_WvsF[region_WvsF$padj_aff_lower< 0.1 ]<- "TRUE"

```


```{r}
# DEG unaff.ECA_WvsF
region_WvsF$sig_unaff.ECA_WvsF<- "FALSE"
region_WvsF$sig_unaff.ECA_WvsF[region_WvsF$padj_unaff_ECA< 0.1 ]<- "TRUE"

# DEG aff.ECA_WvsF
region_WvsF$sig_aff.ECA_WvsF<- "FALSE"
region_WvsF$sig_aff.ECA_WvsF[region_WvsF$padj_aff_ECA< 0.1 ]<- "TRUE"


# DEG unaff.lower_WvsF
region_WvsF$sig_unaff.lower_WvsF<- "FALSE"
region_WvsF$sig_unaff.lower_WvsF[region_WvsF$padj_unaff_lower< 0.1 ]<- "TRUE"

# DEG aff.lower_WvsF
region_WvsF$sig_aff.lower_WvsF<- "FALSE"
region_WvsF$sig_aff.lower_WvsF[region_WvsF$padj_aff_lower< 0.1 ]<- "TRUE"
```

```{r}
region_WvsF[, 14:20] <- lapply(region_WvsF[, 14:20], as.logical)

# Unique DEG in unaffected --> 28 genes
region_WvsF$uniq_sig_unaff.region_WvsF <- with(region_WvsF, 
                                               sig_unaff.ECA_WvsF & 
                                               sig_unaff.lower_WvsF & 
                                               !sig_aff.ECA_WvsF & 
                                               !sig_aff.lower_WvsF)
sum(region_WvsF$uniq_sig_unaff.region_WvsF, na.rm = TRUE)

# Unique DEG in affected --> 460 genes
region_WvsF$uniq_sig_aff.region_WvsF <- with(region_WvsF, 
                                               !sig_unaff.ECA_WvsF & 
                                               !sig_unaff.lower_WvsF & 
                                                sig_aff.ECA_WvsF & 
                                                sig_aff.lower_WvsF)
sum(region_WvsF$uniq_sig_aff.region_WvsF, na.rm = TRUE)

```

```{r}
# DEG common affected region WvsF -->1548
region_WvsF$com_sig_aff.region_WvsF <- with(region_WvsF,
                                                sig_aff.ECA_WvsF & 
                                                sig_aff.lower_WvsF)
cat("Regional common DEG in SCTLD affected corals btw WvsF", sum(region_WvsF$com_sig_aff.region_WvsF, na.rm = TRUE) )

# DEG common unaffected region WvsF-->426
region_WvsF$com_sig_unaff.region_WvsF <- with(region_WvsF,
                                                sig_unaff.ECA_WvsF & 
                                                sig_unaff.lower_WvsF)
cat("Regional common DEG in SCTLD unaffected corals btw WvsF", sum(region_WvsF$com_sig_unaff.region_WvsF, na.rm = TRUE) )

# DEG common unaffected  affected ECA WvsF-->472
region_WvsF$com_sig_ECA.aff.unaff_WvsF <- with(region_WvsF,
                                                sig_unaff.ECA_WvsF & 
                                                sig_aff.ECA_WvsF)
cat("ECA common DEG in SCTLD unaffected and affected corals btw WvsF", sum(region_WvsF$com_sig_ECA.aff.unaff_WvsF, na.rm = TRUE) )

# DEG common unaffected  affected lower WvsF-->1541
region_WvsF$com_sig_lower.aff.unaff_WvsF <- with(region_WvsF,
                                                sig_unaff.lower_WvsF & 
                                                sig_aff.lower_WvsF)
cat("Lower Keys common DEG in SCTLD unaffected and affected corals btw WvsF", sum(region_WvsF$com_sig_lower.aff.unaff_WvsF, na.rm = TRUE) )
```


```{r}
# DEG common unaffected  affected region WvsF-->339
region_WvsF$com_sig_region.aff.unaff_WvsF <- with(region_WvsF,
                                                com_sig_aff.region_WvsF & 
                                                com_sig_unaff.region_WvsF)
cat("Regional common DEG in SCTLD unaffected and affected corals btw WvsF",sum(region_WvsF$com_sig_region.aff.unaff_WvsF, na.rm = TRUE))

```

```{r}
# DEG common unaffected  affected and region WvsF-->339
region_WvsF$com.com_sig_WvsF <- with(region_WvsF,
                                                com_sig_aff.region_WvsF & 
                                                com_sig_unaff.region_WvsF &
                                                com_sig_ECA.aff.unaff_WvsF &
                                                com_sig_lower.aff.unaff_WvsF)
sum(region_WvsF$com.com_sig_WvsF, na.rm = TRUE)

# DEG all common unaffected  affected WvsF-->2162
region_WvsF$all.com_sig_WvsF <- with(region_WvsF,
                                                com_sig_aff.region_WvsF | 
                                                com_sig_unaff.region_WvsF |
                                                com_sig_ECA.aff.unaff_WvsF |
                                               com_sig_lower.aff.unaff_WvsF)
sum(region_WvsF$all.com_sig_WvsF, na.rm = TRUE)

```


```{r}
# Coral ECA gene direction per susceptibility
region_WvsF$DEG_ECA_aff.WvsF<- "NO"
region_WvsF$DEG_ECA_aff.WvsF[region_WvsF$log_aff_ECA > 0 & region_WvsF$padj_aff_ECA < 0.1] <- "UP"
region_WvsF$DEG_ECA_aff.WvsF[region_WvsF$log_aff_ECA < 0 & region_WvsF$padj_aff_ECA < 0.1] <- "DOWN"

region_WvsF$DEG_ECA_unaff.WvsF<- "NO"
region_WvsF$DEG_ECA_unaff.WvsF[region_WvsF$log_unaff_ECA > 0 & region_WvsF$padj_unaff_ECA < 0.1] <- "UP"
region_WvsF$DEG_ECA_unaff.WvsF[region_WvsF$log_unaff_ECA < 0 & region_WvsF$padj_unaff_ECA < 0.1] <- "DOWN"


# Lower Keys gene direction per susceptibility
region_WvsF$DEG_lower_aff.WvsF<- "NO"
region_WvsF$DEG_lower_aff.WvsF[region_WvsF$log_aff_lower > 0 & region_WvsF$padj_aff_lower < 0.1] <- "UP"
region_WvsF$DEG_lower_aff.WvsF[region_WvsF$log_aff_lower < 0 & region_WvsF$padj_aff_lower < 0.1] <- "DOWN"

region_WvsF$DEG_lower_unaff.WvsF<- "NO"
region_WvsF$DEG_lower_unaff.WvsF[region_WvsF$log_unaff_lower > 0 & region_WvsF$padj_unaff_lower < 0.1] <- "UP"
region_WvsF$DEG_lower_unaff.WvsF[region_WvsF$log_unaff_lower < 0 & region_WvsF$padj_unaff_lower < 0.1] <- "DOWN"
```

```{r}
region_WvsF_anno<- region_WvsF %>% left_join(anno, by="ofav_gene")
```

```{r}
region_plot_long <- region_WvsF %>%
  select(com_sig_unaff.region_WvsF, com_sig_aff.region_WvsF,com_sig_lower.aff.unaff_WvsF,com_sig_ECA.aff.unaff_WvsF,all.com_sig_WvsF,uniq_sig_unaff.region_WvsF,uniq_sig_aff.region_WvsF) %>%
  pivot_longer(cols = everything(), names_to = "DEG_category", values_to = "Value") %>%
  filter(Value == TRUE)

# Plot the counts of TRUE for each region
ggplot(region_plot_long, aes(x = DEG_category, fill=DEG_category)) +
  geom_bar(stat = "count", position = "dodge") +
  scale_fill_viridis_d(option = "D") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) 
  labs(
    title = "Counts of Regional common DEG of affected and unaffected corals",
    x = "DEG_category",
    y = "Count"
  ) +
  theme_minimal()
```


#### Extracting DEG Winter vs Fall 
##### DEG Winter vs Fall per region and susceptibility
```{r}
DEG_ECA_aff.WvsF<- region_WvsF_anno %>% filter(padj_aff_ECA < 0.1)

DEG_ECA_unaff.WvsF<- region_WvsF_anno %>% filter(padj_unaff_ECA < 0.1)

DEG_lower_aff.WvsF<- region_WvsF_anno %>% filter(padj_aff_lower < 0.1)

DEG_lower_unaff.WvsF<- region_WvsF_anno %>% filter(padj_unaff_lower < 0.1)

```

##### DEG Winter vs Fall common and unique
```{r}
# All DEG --> 3419
DEG.all_WvsF<- region_WvsF_anno %>% filter(sig_genes.WvsF == TRUE)

# all_sig_genes.WvsF<- all_sig_genes.WvsF %>% select(-gDNA,-mRNA, -"CDS-transcript",-Translation)

# DEG  btw region and unaffected corals -->1499
DEG.all.unaff_WvsF<- region_WvsF_anno %>% filter(sig_unaff.region_WvsF == TRUE)

# DEG  btw region and affected corals -->3161
DEG.all.aff_WvsF<- region_WvsF_anno %>% filter(sig_aff.region_WvsF == TRUE)

# DEG uniq btw region and affected corals --> 308
DEG.uniq.aff_WvsF<- region_WvsF_anno %>% filter(uniq_sig_aff.region_WvsF == TRUE)

# DEG uniq btw region and unaffected corals -->3
DEG.uniq.unaff_WvsF<- region_WvsF_anno %>% filter(uniq_sig_unaff.region_WvsF == TRUE)

# DEG common btw region and unaffected corals --> 994
DEG.com.aff_WvsF<- region_WvsF_anno %>% filter(com_sig_aff.region_WvsF == TRUE)

# DEG common btw region and affected corals -->220
DEG.com.unaff_WvsF<- region_WvsF_anno %>% filter(com_sig_unaff.region_WvsF == TRUE)

# DEG all common -->190
DEG.all.com_WvsF<- region_WvsF_anno %>% filter(all.com_sig_WvsF == TRUE)


# DEG common amongst all contrast
DEG.com.com_WvsF<- region_WvsF_anno %>% filter(com.com_sig_WvsF == TRUE)
```


#### Counts per contrast and all sig genes
```{r}
# VST counts for affected for both regions WvsF

ids_regions.aff<- DEG.all.aff_WvsF$ofav_gene

count_regions.aff.WvsF_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_regions.aff),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_regions.aff.WvsF_vst))
all(rownames(colData_4M) == colnames(count_regions.aff.WvsF_vst))


# VST counts for DEG common in both regions in affected: WvsF

ids_com.region.aff<- DEG.com.aff_WvsF$ofav_gene

count_com.region.aff.WvsF_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_com.region.aff),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_com.region.aff.WvsF_vst))
all(rownames(colData_4M) == colnames(count_com.region.aff.WvsF_vst))

# VST counts for DEG common in both regions in unaffected: WvsF

ids_com.region.unaff<- DEG.com.unaff_WvsF$ofav_gene

count_com.region.unaff.WvsF_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_com.region.unaff),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_com.region.unaff.WvsF_vst))
all(rownames(colData_4M) == colnames(count_com.region.unaff.WvsF_vst))


# VST counts for uniq unaffected both regions

ids_uniq.region.unaff<- DEG.uniq.unaff_WvsF$ofav_gene

count_uniq.region.unaff.WvsF_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_uniq.region.unaff),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_uniq.region.unaff.WvsF_vst))
all(rownames(colData_4M) == colnames(count_uniq.region.unaff.WvsF_vst))

# VST counts for uniq affected both regions

ids_uniq.region.aff<- DEG.uniq.aff_WvsF$ofav_gene

count_uniq.region.aff.WvsF_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_uniq.region.aff),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_uniq.region.aff.WvsF_vst))
all(rownames(colData_4M) == colnames(count_uniq.region.aff.WvsF_vst))


# VST counts for all common seasonal change

ids_all.com_WvsF<- DEG.all.com_WvsF$ofav_gene

count_all.com_WvsF_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_all.com_WvsF),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_all.com_WvsF_vst))
all(rownames(colData_4M) == colnames(count_all.com_WvsF_vst))


# VST counts for  common across contrast

ids_com.com_WvsF<- DEG.com.com_WvsF$ofav_gene

count_com.com_WvsF_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_com.com_WvsF),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_com.com_WvsF_vst))
all(rownames(colData_4M) == colnames(count_com.com_WvsF_vst))


# VST counts for all sig genes from the 4 seasonal contrast

ids_all.sig_WvsF<- DEG.all_WvsF$ofav_gene

count_all.sig_WvsF_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_all.sig_WvsF),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_all.sig_WvsF_vst))
all(rownames(colData_4M) == colnames(count_all.sig_WvsF_vst))
```

#### Plots seasonal effect in gene expression per susceptability and region

##### PCA plots
###### PCA all sign genes  WvsF
```{r}
pcaData = plotPCA(count_com.region.unaff.WvsF_vst, intgroup=c("colnames",'SCTLD_RNA',"Region","sample_month","time_point","Location","SCTLD_RNA_region","SCTLD_RNA_SP_region", "sample_day","percent_uniq","Sample_ID","SCTLD_RNA_SP","SP_region","Core_Site_Location","Histopath_supported_resistance","SCTLD_RNA3","sample_period","Plate_Number","Disease_on_colony_when_coring","SCTLD_RNA_Region_season"), 
returnData=TRUE)

pcaData$SCTLD_RNA<- as.factor(pcaData$SCTLD_RNA)
pcaData$SCTLD_RNA<- factor(pcaData$SCTLD_RNA,levels=c("SCTLD_affected","SCTLD_unaffected"))

pcaData$SP_region<- as.factor(pcaData$SP_region)
pcaData$SP_region<- factor(pcaData$SP_region,levels=c("SP2_ECA","SP2_Lower","SP3_ECA","SP3_Lower"))
                                

pcaData$SCTLD_RNA_region<- as.factor(pcaData$SCTLD_RNA_region)
pcaData$SCTLD_RNA_region<- factor(pcaData$SCTLD_RNA_region,levels=c("SCTLD_affected_ECA","SCTLD_unaffected_ECA","SCTLD_affected_Lower","SCTLD_unaffected_Lower"))

pcaData$SCTLD_RNA_SP_region<- as.factor(pcaData$SCTLD_RNA_SP_region)
pcaData$SCTLD_RNA_SP_region<- factor(pcaData$SCTLD_RNA_SP_region,levels=c("SCTLD_affected_SP3_ECA","SCTLD_unaffected_SP3_ECA","SCTLD_affected_SP3_Lower","SCTLD_unaffected_SP3_Lower","SCTLD_affected_SP2_ECA","SCTLD_unaffected_SP2_ECA","SCTLD_affected_SP2_Lower","SCTLD_unaffected_SP2_Lower"))

pcaData$SCTLD_RNA_Region_season<- factor(pcaData$SCTLD_RNA_Region_season,levels=c("SCTLD_affected_ECA_winter","SCTLD_unaffected_ECA_winter","SCTLD_affected_Lower_winter","SCTLD_unaffected_Lower_winter","SCTLD_affected_Lower_fall","SCTLD_unaffected_Lower_fall","SCTLD_affected_ECA_fall","SCTLD_unaffected_ECA_fall"))

percentVar = round(100 * attr(pcaData, "percentVar"))

# class_colors <- c("SCTLD_affected" ="#009E73",  # Teal
#                    "SCTLD_unaffected" = "#E69F00")
# 
# class_colors <- c("SP2_ECA" ="#F2994A" ,  # orange
#                    "SP2_Lower" = "#D96C20",  # Light Blue
#                    "SP3_ECA" = "#2D9CDB",  # Teal
#                    "SP3_Lower" = "#1B73A5")  # green


class_colors <- c("SCTLD_affected_ECA_winter" ="#D3D3D3" ,  
                   "SCTLD_unaffected_ECA_winter" = "#2D9CDB",  
                   "SCTLD_affected_ECA_fall" = "#D3D3D3",  
                   "SCTLD_unaffected_ECA_fall" = "#E57333",
                   "SCTLD_affected_Lower_winter" = "#D3D3D3",  
                   "SCTLD_unaffected_Lower_winter" = "#1E6F9F",
                   "SCTLD_affected_Lower_fall" = "#D3D3D3",  
                   "SCTLD_unaffected_Lower_fall" = "#F7C948")  

# class_colors <- c("1" ="#009E73",  # Teal
#                    "0" = "#E69F00")

# png("results/8_mint/PCA_mint_comp1.png", width=10, height=10, units = "in", res = 300)
ggplot(pcaData, aes(PC1, PC2, colour = SCTLD_RNA_Region_season, shape = SCTLD_RNA)) + 
geom_point(size = 3) + 
stat_ellipse(aes(group = SCTLD_RNA_Region_season),type = "t", linetype = "dashed", level = 0.95) +
# geom_text_repel(aes(label = Sample_ID), nudge_x = -1, nudge_y = 0.2, size = 3) +
scale_color_manual(values = class_colors) +
ggtitle("PCA DESeq2:vst filtered, DEG WvsF in affected corals ~ SCTLD_RNA_Region_season") +
xlab(paste0("PC1: ",percentVar[1],"% variance")) +
ylab(paste0("PC2: ",percentVar[2],"% variance")) +
theme_bw() + 
theme_minimal()
# theme(panel.background = element_rect(fill = "lightgrey"),legend.position = "top")
# dev.off()

beep()
```
###### PCA sign genes common affected WvsF
```{r}
pcaData = plotPCA(count_com.region.aff.WvsF_vst, intgroup=c("colnames",'SCTLD_RNA',"Region","sample_month","time_point","Location","SCTLD_RNA_region","SCTLD_RNA_SP_region", "sample_day","percent_uniq","Sample_ID","SCTLD_RNA_SP","SP_region","Core_Site_Location","Histopath_supported_resistance","SCTLD_RNA3","sample_period","Plate_Number","Disease_on_colony_when_coring","SCTLD_RNA_Region_season"), 
returnData=TRUE)

pcaData$SCTLD_RNA<- as.factor(pcaData$SCTLD_RNA)
pcaData$SCTLD_RNA<- factor(pcaData$SCTLD_RNA,levels=c("SCTLD_affected","SCTLD_unaffected"))

pcaData$SP_region<- as.factor(pcaData$SP_region)
pcaData$SP_region<- factor(pcaData$SP_region,levels=c("SP2_ECA","SP2_Lower","SP3_ECA","SP3_Lower"))
                                

pcaData$SCTLD_RNA_region<- as.factor(pcaData$SCTLD_RNA_region)
pcaData$SCTLD_RNA_region<- factor(pcaData$SCTLD_RNA_region,levels=c("SCTLD_affected_ECA","SCTLD_unaffected_ECA","SCTLD_affected_Lower","SCTLD_unaffected_Lower"))

pcaData$SCTLD_RNA_SP_region<- as.factor(pcaData$SCTLD_RNA_SP_region)
pcaData$SCTLD_RNA_SP_region<- factor(pcaData$SCTLD_RNA_SP_region,levels=c("SCTLD_affected_SP3_ECA","SCTLD_unaffected_SP3_ECA","SCTLD_affected_SP3_Lower","SCTLD_unaffected_SP3_Lower","SCTLD_affected_SP2_ECA","SCTLD_unaffected_SP2_ECA","SCTLD_affected_SP2_Lower","SCTLD_unaffected_SP2_Lower"))

pcaData$SCTLD_RNA_Region_season<- factor(pcaData$SCTLD_RNA_Region_season,levels=c("SCTLD_affected_ECA_winter","SCTLD_unaffected_ECA_winter","SCTLD_affected_Lower_winter","SCTLD_unaffected_Lower_winter","SCTLD_affected_Lower_fall","SCTLD_unaffected_Lower_fall","SCTLD_affected_ECA_fall","SCTLD_unaffected_ECA_fall"))

percentVar = round(100 * attr(pcaData, "percentVar"))

# class_colors <- c("SCTLD_affected" ="#009E73",  # Teal
#                    "SCTLD_unaffected" = "#E69F00")
# 
# class_colors <- c("SP2_ECA" ="#F2994A" ,  # Grey
#                    "SP2_Lower" = "#D96C20",  # Light Blue
#                    "SP3_ECA" = "#2D9CDB",  # Teal
#                    "SP3_Lower" = "#1B73A5")  # Orange


class_colors <- c("SCTLD_affected_ECA_winter" ="#A37ACC" ,  
                   "SCTLD_unaffected_ECA_winter" = "#D3D3D3",  
                   "SCTLD_affected_ECA_fall" = "#6BBF59",  
                   "SCTLD_unaffected_ECA_fall" = "#D3D3D3",
                   "SCTLD_affected_Lower_winter" = "#5E4B9E",  
                   "SCTLD_unaffected_Lower_winter" = "#D3D3D3",
                   "SCTLD_affected_Lower_fall" = "#3B7F4C",  
                   "SCTLD_unaffected_Lower_fall" = "#D3D3D3")

# png("results/8_mint/PCA_mint_comp1.png", width=10, height=10, units = "in", res = 300)
ggplot(pcaData, aes(PC1, PC2, colour = SCTLD_RNA_Region_season, shape = SCTLD_RNA)) + 
geom_point(size = 3) + 
stat_ellipse(aes(group = SCTLD_RNA_Region_season),type = "t", linetype = "dashed", level = 0.95) +
# geom_text_repel(aes(label = Sample_ID), nudge_x = -1, nudge_y = 0.2, size = 3) +
scale_color_manual(values = class_colors) +
ggtitle("PCA DESeq2:vst filtered, DEG WvsF in affected corals ~ SCTLD_RNA_Region_season") +
xlab(paste0("PC1: ",percentVar[1],"% variance")) +
ylab(paste0("PC2: ",percentVar[2],"% variance")) +
theme_bw() + 
theme_minimal()
# theme(panel.background = element_rect(fill = "lightgrey"),legend.position = "top")
# dev.off()

beep()
```


##### Heatmaps for sesonal effect per susceptability and region

###### Common region SCTLD affected
```{r}
gene_aff_region.WvsF<- DEG.com.aff_WvsF
rownames(gene_aff_region.WvsF)<- gene_aff_region.WvsF$ofav_gene
# gene_aff_region.WvsF<- gene_aff_region.WvsF[order(gene_aff_region.WvsF$log_aff_region.WvsF, decreasing = TRUE),]

aff_region.WvsF_df<- colData_4M %>% filter(SCTLD_RNA == "SCTLD_affected")
samples_aff_region.WvsF<- aff_region.WvsF_df$Sample_ID


vst_out_aff_region.WvsF<-count_com.region.aff.WvsF_vst[(rownames(count_com.region.aff.WvsF_vst) %in% gene_aff_region.WvsF),]

vst_out_aff_region.WvsF<-count_com.region.aff.WvsF_vst[,(colnames(count_com.region.aff.WvsF_vst) %in% samples_aff_region.WvsF)]

# making the rownames and column names identical
all(rownames(samples_aff_region.WvsF) %in% colnames(vst_out_aff_region.WvsF))
all(rownames(samples_aff_region.WvsF) == colnames(vst_out_aff_region.WvsF))

mat_aff_region.WvsF<-assay(vst_out_aff_region.WvsF) #sig genes x samples
mat_aff_region.WvsF.scaled <- t(apply(mat_aff_region.WvsF, 1, scale)) #center and scale each column (Z-score) then transpose
colnames(mat_aff_region.WvsF.scaled)<-colnames(mat_aff_region.WvsF)

aff.WvsF <- as.factor(aff_region.WvsF_df$SCTLD_RNA)
aff.WvsF <- factor(aff.WvsF,levels = c("SCTLD_affected"))

aff_region.WvsF <- as.factor(aff_region.WvsF_df$Region)
aff_region.WvsF <- factor(aff_region.WvsF,levels = c("ECA","Lower"))

SP_aff.WvsF <- as.factor(aff_region.WvsF_df$sample_period)
SP_aff.WvsF <- factor(SP_aff.WvsF,levels = c("SP2","SP3"))


ha_aff_region.WvsF <- HeatmapAnnotation( 
                         Season  = SP_aff.WvsF,
                         Region = aff_region.WvsF,
                         SCTLD_affected = aff.WvsF,
                         col = list(
                           Season = c("SP2"= "#D55E00","SP3"="#0072B2"),
                           Region = c("ECA"="#009E73","Lower"="#F0E442"),
                           SCTLD_affected = c("SCTLD_affected" = "#E69F00")
                           ),
                         annotation_legend_param = list(
                           Season = list(title = "Season"),
                           Region = list(title = "Region"),
                           SCTLD_affected = list(title = "SCTLD susceptibility")),
                           show_legend = c(Season = TRUE, Region = TRUE, SCTLD = TRUE) 
                         )
  # sample_data = anno_barplot(mat_aff_region.WvsF, gp = gpar(fill = "blue"), height = unit(2, "cm"))  # Barplot for samples

```

```{r}
Heatmap(mat_aff_region.WvsF.scaled, 
        cluster_columns =  F,  # Disable clustering for columns
        cluster_rows =T,     
        top_annotation = ha_aff_region.WvsF)
```

###### Common region SCTLD unaffected
```{r}
gene_unaff_region.WvsF<- DEG.com.unaff_WvsF
rownames(gene_unaff_region.WvsF)<- gene_unaff_region.WvsF$ofav_gene
# gene_unaff_region.WvsF<- gene_unaff_region.WvsF[order(gene_unaff_region.WvsF$log_unaff_region.WvsF, decreasing = TRUE),]

unaff_region.WvsF_df<- colData_4M %>% filter(SCTLD_RNA == "SCTLD_unaffected")
samples_unaff_region.WvsF<- unaff_region.WvsF_df$Sample_ID


vst_out_unaff_region.WvsF<-count_com.region.unaff.WvsF_vst[(rownames(count_com.region.unaff.WvsF_vst) %in% gene_unaff_region.WvsF),]

vst_out_unaff_region.WvsF<-count_com.region.unaff.WvsF_vst[,(colnames(count_com.region.unaff.WvsF_vst) %in% samples_unaff_region.WvsF)]

# making the rownames and column names identical
all(rownames(samples_unaff_region.WvsF) %in% colnames(vst_out_unaff_region.WvsF))
all(rownames(samples_unaff_region.WvsF) == colnames(vst_out_unaff_region.WvsF))

mat_unaff_region.WvsF<-assay(vst_out_unaff_region.WvsF) #sig genes x samples
mat_unaff_region.WvsF.scaled <- t(apply(mat_unaff_region.WvsF, 1, scale)) #center and scale each column (Z-score) then transpose
colnames(mat_unaff_region.WvsF.scaled)<-colnames(mat_unaff_region.WvsF)

unaff.WvsF <- as.factor(unaff_region.WvsF_df$SCTLD_RNA)
unaff.WvsF <- factor(unaff.WvsF,levels = c("SCTLD_unaffected"))

unaff_region.WvsF <- as.factor(unaff_region.WvsF_df$Region)
unaff_region.WvsF <- factor(unaff_region.WvsF,levels = c("ECA","Lower"))

SP_unaff.WvsF <- as.factor(unaff_region.WvsF_df$sample_period)
SP_unaff.WvsF <- factor(SP_unaff.WvsF,levels = c("SP2","SP3"))


ha_unaff_region.WvsF <- HeatmapAnnotation( 
                         Season  = SP_unaff.WvsF,
                         Region = unaff_region.WvsF,
                         SCTLD = unaff.WvsF,
                         col = list(
                           Season = c("SP2"= "#D55E00","SP3"="#0072B2"),
                           Region = c("ECA"="#009E73","Lower"="#F0E442"),
                           SCTLD = c("SCTLD_unaffected" = "#56B4E9")
                           ),
                         annotation_legend_param = list(
                           Season = list(title = "Season"),
                           Region = list(title = "Region"),
                           SCTLD = list(title = "SCTLD susceptibility")),
                           show_legend = c(Season = TRUE, Region = TRUE, SCTLD = FALSE) 
                         )
  # sample_data = anno_barplot(mat_unaff_region.WvsF, gp = gpar(fill = "blue"), height = unit(2, "cm"))  # Barplot for samples

```

```{r}
Heatmap(mat_unaff_region.WvsF.scaled, 
        cluster_columns =  T,  # Disable clustering for columns
        cluster_rows =T,     
        top_annotation = ha_unaff_region.WvsF)
```


###### Common for all contrast all samples
```{r}
gene_com.com.WvsF<- DEG.com.com_WvsF
rownames(gene_com.com.WvsF)<- gene_com.com.WvsF$ofav_gene
# gene_com.com.WvsF<- gene_com.com.WvsF[order(gene_com.com.WvsF$log_com.com.WvsF, decreasing = TRUE),]

com.com.WvsF_df<- colData_4M 
samples_com.com.WvsF<- com.com.WvsF_df$Sample_ID


vst_out_com.com.WvsF<-count_com.com_WvsF_vst[(rownames(count_com.com_WvsF_vst) %in% gene_com.com.WvsF),]

vst_out_com.com.WvsF<-count_com.com_WvsF_vst[,(colnames(count_com.com_WvsF_vst) %in% samples_com.com.WvsF)]

# making the rownames and column names identical
all(rownames(samples_com.com.WvsF) %in% colnames(vst_out_com.com.WvsF))
all(rownames(samples_com.com.WvsF) == colnames(vst_out_com.com.WvsF))

mat_com.com.WvsF<-assay(vst_out_com.com.WvsF) #sig genes x samples
mat_com.com.WvsF.scaled <- t(apply(mat_com.com.WvsF, 1, scale)) #center and scale each column (Z-score) then transpose
colnames(mat_com.com.WvsF.scaled)<-colnames(mat_com.com.WvsF)

sctld.WvsF <- as.factor(com.com.WvsF_df$SCTLD_RNA)
sctld.WvsF <- factor(sctld.WvsF,levels = c("SCTLD_unaffected","SCTLD_affected"))

region.WvsF <- as.factor(com.com.WvsF_df$Region)
region.WvsF <- factor(region.WvsF,levels = c("ECA","Lower"))

SP.WvsF <- as.factor(com.com.WvsF_df$sample_period)
SP.WvsF <- factor(SP.WvsF,levels = c("SP2","SP3"))


ha_com.com.WvsF <- HeatmapAnnotation( 
                         Season  = SP.WvsF,
                         Region = region.WvsF,
                         SCTLD = sctld.WvsF,
                         col = list(
                           Season = c("SP2"= "#D55E00","SP3"="#0072B2"),
                           Region = c("ECA"="#009E73","Lower"="#F0E442"),
                           SCTLD = c("SCTLD_unaffected" ="#56B4E9","SCTLD_affected" = "#E69F00")
                           ),
                         annotation_legend_param = list(
                           Season = list(title = "Season"),
                           Region = list(title = "Region"),
                           SCTLD = list(title = "SCTLD susceptibility")),
                           show_legend = c(Season = TRUE, Region = TRUE, SCTLD = TRUE) 
                         )
  # sample_data = anno_barplot(mat_com.com.WvsF, gp = gpar(fill = "blue"), height = unit(2, "cm"))  # Barplot for samples

```

```{r}
Heatmap(mat_com.com.WvsF.scaled, 
        cluster_columns =  F,  # Disable clustering for columns
        cluster_rows =T,     
        top_annotation = ha_com.com.WvsF)
```




###### Common for all contrast only affected samples
```{r}
gene_com.com_aff_WvsF<- DEG.com.com_WvsF
rownames(gene_com.com_aff_WvsF)<- gene_com.com_aff_WvsF$ofav_gene
# gene_com.com_aff_WvsF<- gene_com.com_aff_WvsF[order(gene_com.com_aff_WvsF$log_com.com.WvsF, decreasing = TRUE),]

com.com_aff_WvsF_df<- colData_4M %>% filter(SCTLD_RNA == "SCTLD_affected")
samples_com.com_aff_WvsF<- com.com_aff_WvsF_df$Sample_ID


vst_out_com.com_aff_WvsF<-count_com.com_WvsF_vst[(rownames(count_com.com_WvsF_vst) %in% gene_com.com_aff_WvsF),]

vst_out_com.com_aff_WvsF<-count_com.com_WvsF_vst[,(colnames(count_com.com_WvsF_vst) %in% samples_com.com_aff_WvsF)]

# making the rownames and column names identical
all(rownames(samples_com.com_aff_WvsF) %in% colnames(vst_out_com.com_aff_WvsF))
all(rownames(samples_com.com_aff_WvsF) == colnames(vst_out_com.com_aff_WvsF))

mat_com.com_aff_WvsF<-assay(vst_out_com.com_aff_WvsF) #sig genes x samples
mat_com.com_aff_WvsF.scaled <- t(apply(mat_com.com_aff_WvsF, 1, scale)) #center and scale each column (Z-score) then transpose
colnames(mat_com.com_aff_WvsF.scaled)<-colnames(mat_com.com_aff_WvsF)

sctld.aff.WvsF <- as.factor(com.com_aff_WvsF_df$SCTLD_RNA)
sctld.aff.WvsF <- factor(sctld.aff.WvsF,levels = c("SCTLD_affected"))

region.WvsF <- as.factor(com.com_aff_WvsF_df$Region)
region.WvsF <- factor(region.WvsF,levels = c("ECA","Lower"))

SP.WvsF <- as.factor(com.com_aff_WvsF_df$sample_period)
SP.WvsF <- factor(SP.WvsF,levels = c("SP2","SP3"))


ha_com.com_aff_WvsF <- HeatmapAnnotation( 
                         Season  = SP.WvsF,
                         Region = region.WvsF,
                         SCTLD = sctld.aff.WvsF,
                         col = list(
                           Season = c("SP2"= "#D55E00","SP3"="#0072B2"),
                           Region = c("ECA"="#009E73","Lower"="#F0E442"),
                           SCTLD = c("SCTLD_affected" = "#E69F00")
                           ),
                         annotation_legend_param = list(
                           Season = list(title = "Season"),
                           Region = list(title = "Region"),
                           SCTLD = list(title = "SCTLD susceptibility")),
                           show_legend = c(Season = TRUE, Region = TRUE, SCTLD = TRUE) 
                         )
  # sample_data = anno_barplot(mat_com.com_aff_WvsF, gp = gpar(fill = "blue"), height = unit(2, "cm"))  # Barplot for samples

```

```{r}
Heatmap(mat_com.com_aff_WvsF.scaled, 
        cluster_columns =  F,  # Disable clustering for columns
        cluster_rows =T,     
        top_annotation = ha_com.com_aff_WvsF)
```

###### Common for all susceptibility contrast only unaffected samples
```{r}
gene_com.com_unaff_WvsF<- DEG.com.com_WvsF
rownames(gene_com.com_unaff_WvsF)<- gene_com.com_unaff_WvsF$ofav_gene
# gene_com.com_unaff_WvsF<- gene_com.com_unaff_WvsF[order(gene_com.com_unaff_WvsF$log_com.com.WvsF, decreasing = TRUE),]

com.com_unaff_WvsF_df<- colData_4M %>% filter(SCTLD_RNA == "SCTLD_unaffected")
samples_com.com_unaff_WvsF<- com.com_unaff_WvsF_df$Sample_ID


vst_out_com.com_unaff_WvsF<-count_com.com_WvsF_vst[(rownames(count_com.com_WvsF_vst) %in% gene_com.com_unaff_WvsF),]

vst_out_com.com_unaff_WvsF<-count_com.com_WvsF_vst[,(colnames(count_com.com_WvsF_vst) %in% samples_com.com_unaff_WvsF)]

# making the rownames and column names identical
all(rownames(samples_com.com_unaff_WvsF) %in% colnames(vst_out_com.com_unaff_WvsF))
all(rownames(samples_com.com_unaff_WvsF) == colnames(vst_out_com.com_unaff_WvsF))

mat_com.com_unaff_WvsF<-assay(vst_out_com.com_unaff_WvsF) #sig genes x samples
mat_com.com_unaff_WvsF.scaled <- t(apply(mat_com.com_unaff_WvsF, 1, scale)) #center and scale each column (Z-score) then transpose
colnames(mat_com.com_unaff_WvsF.scaled)<-colnames(mat_com.com_unaff_WvsF)

sctld.unaff.WvsF <- as.factor(com.com_unaff_WvsF_df$SCTLD_RNA)
sctld.unaff.WvsF <- factor(sctld.unaff.WvsF,levels = c("SCTLD_unaffected"))

region.WvsF <- as.factor(com.com_unaff_WvsF_df$Region)
region.WvsF <- factor(region.WvsF,levels = c("ECA","Lower"))

SP.WvsF <- as.factor(com.com_unaff_WvsF_df$sample_period)
SP.WvsF <- factor(SP.WvsF,levels = c("SP2","SP3"))


ha_com.com_unaff_WvsF <- HeatmapAnnotation( 
                         Season  = SP.WvsF,
                         Region = region.WvsF,
                         SCTLD = sctld.unaff.WvsF,
                         col = list(
                           Season = c("SP2"= "#D55E00","SP3"="#0072B2"),
                           Region = c("ECA"="#009E73","Lower"="#F0E442"),
                           SCTLD = c("SCTLD_unaffected" = "#56B4E9")
                           ),
                         annotation_legend_param = list(
                           Season = list(title = "Season"),
                           Region = list(title = "Region"),
                           SCTLD = list(title = "SCTLD susceptibility")),
                           show_legend = c(Season = TRUE, Region = TRUE, SCTLD = TRUE) 
                         )
  # sample_data = anno_barplot(mat_com.com_unaff_WvsF, gp = gpar(fill = "blue"), height = unit(2, "cm"))  # Barplot for samples

```

```{r}
Heatmap(mat_com.com_unaff_WvsF.scaled, 
        cluster_columns =  T,  # Disable clustering for columns
        cluster_rows =T,     
        top_annotation = ha_com.com_unaff_WvsF)
```






#### Full transcriptome metadata for 8 contrast
```{r}
trans_all<- suscept_season_region %>% full_join(region_WvsF, by="ofav_gene")

```

#### List of DEG from 8 contrast
```{r}
dim(DEG.all_WvsF) #--> 4368
dim(sig_suscept_season_region)#-->121
DEG.all_aff.vs.unaff<- sig_suscept_season_region[,c(1:22)]

all_contrast_DEG<- DEG.all_WvsF %>% full_join(DEG.all_aff.vs.unaff,by="ofav_gene")

dim(all_contrast_DEG)
```

```{r}
# VST counts for all sig genes from the 8  contrast

ids_all_contrast_DEG<- all_contrast_DEG$ofav_gene

count_all_contrast_DEG_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_all_contrast_DEG),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_all_contrast_DEG_vst))
all(rownames(colData_4M) == colnames(count_all_contrast_DEG_vst))
```


########### KEEP EDITING FROM HERE ##########


#### Contrast combining factors
```{r}
resultsNames(dds_sctld.reg.season)

```

##### DEG btw SCTLD_affected in fall vs SCTLD_unaffected in fall--> for both regions

```{r}
res_both_affvsunaff_fall<- results(dds_sctld.reg.season, contrast = list(
    c("SCTLD_RNA_Region_seasonSCTLD_affected_ECA_fall", "SCTLD_RNA_Region_seasonSCTLD_affected_Lower_fall"),
    c("SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_fall", "SCTLD_RNA_Region_seasonSCTLD_unaffected_Lower_fall")
  ),
  listValues = c(0.5, -0.5) )

resOrdered_both_affvsunaff_fall<- as.data.frame(
               
                                 res_both_affvsunaff_fall[order(res_both_affvsunaff_fall$padj),])
                
DEG_both_affvsunaff_fall <-resOrdered_both_affvsunaff_fall %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_both_affvsunaff_fall)) %>%
   
                                filter(padj<0.1)


```

```{r}
res_both_affvsunaff_winter<- results(dds_sctld.reg.season, contrast = list(
    c("SCTLD_RNA_Region_seasonSCTLD_affected_ECA_winter", "SCTLD_RNA_Region_seasonSCTLD_affected_Lower_winter"),
    c("SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_winter", "SCTLD_RNA_Region_seasonSCTLD_unaffected_Lower_winter")
  ),
  listValues = c(0.5, -0.5) )

resOrdered_both_affvsunaff_winter<- as.data.frame(
               
                                 res_both_affvsunaff_winter[order(res_both_affvsunaff_winter$padj),])
                
DEG_both_affvsunaff_winter <-resOrdered_both_affvsunaff_winter %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_both_affvsunaff_winter)) %>%
   
                                filter(padj<0.1)
```

```{r}
res_both_season<- results(dds_sctld.reg.season, contrast = list(
    c("SCTLD_RNA_Region_seasonSCTLD_affected_ECA_winter", "SCTLD_RNA_Region_seasonSCTLD_affected_Lower_winter","SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_winter", "SCTLD_RNA_Region_seasonSCTLD_unaffected_Lower_winter"),
    c("SCTLD_RNA_Region_seasonSCTLD_affected_ECA_fall", "SCTLD_RNA_Region_seasonSCTLD_affected_Lower_fall","SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_fall", "SCTLD_RNA_Region_seasonSCTLD_unaffected_Lower_fall")
  ),
  listValues = c(0.5, -0.5) )

resOrdered_both_season<- as.data.frame(
               
                                 res_both_season[order(res_both_season$padj),])
                
DEG_both_season <-resOrdered_both_season %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_both_season)) %>%
   
                                filter(padj<0.1)
```


```{r}
res_both_season_aff<- results(dds_sctld.reg.season, contrast = list(
    c("SCTLD_RNA_Region_seasonSCTLD_affected_ECA_winter", "SCTLD_RNA_Region_seasonSCTLD_affected_Lower_winter"),
    c("SCTLD_RNA_Region_seasonSCTLD_affected_ECA_fall", "SCTLD_RNA_Region_seasonSCTLD_affected_Lower_fall")
  ),
  listValues = c(0.5, -0.5) )

resOrdered_both_season_aff<- as.data.frame(
               
                                 res_both_season_aff[order(res_both_season_aff$padj),])
                
DEG_both_season_aff <-resOrdered_both_season_aff %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_both_season_aff)) %>%
   
                                filter(padj<0.1)
```

```{r}
res_both_season_unaff<- results(dds_sctld.reg.season, contrast = 
    c("SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_winter", "SCTLD_RNA_Region_seasonSCTLD_unaffected_Lower_winter"),
    c("SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_fall", "SCTLD_RNA_Region_seasonSCTLD_unaffected_Lower_fall"), 
    listValues =c(0.5,-0.5) )

resOrdered_both_season_unaff<- as.data.frame(
               
                                 res_both_season_unaff[order(res_both_season_unaff$padj),])
                
DEG_both_season_unaff <-resOrdered_both_season_unaff %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_both_season_unaff)) %>%
   
                                filter(padj<0.1)
```

```{r}
# Define condition coefficients
unaffected_conditions <- c("SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_fall", 
      "SCTLD_RNA_Region_seasonSCTLD_unaffected_Lower_fall",
      "SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_winter", 
      "SCTLD_RNA_Region_seasonSCTLD_unaffected_Lower_winter")
affected_conditions <- c("SCTLD_RNA_Region_seasonSCTLD_affected_ECA_winter",
      "SCTLD_RNA_Region_seasonSCTLD_affected_Lower_winter",
      "SCTLD_RNA_Region_seasonSCTLD_affected_ECA_fall", 
      "SCTLD_RNA_Region_seasonSCTLD_affected_Lower_fall")

# Create a contrast vector for averaging
contrast_vector <- c(rep(1/length(unaffected_conditions), length(unaffected_conditions)),
                     rep(-1/length(affected_conditions), length(affected_conditions)))

# Perform the contrast
res_affvsunaff <- results(dds_sctld.reg.season, contrast = list(c(unaffected_conditions, affected_conditions)), 
               listValues = c(0.5,-0.5))
```

```{r}
res_affvsunaff<- results(dds_sctld.reg.season, contrast = 
    c("SCTLD_RNA_Region_seasonSCTLD_affected_ECA_winter",
      "SCTLD_RNA_Region_seasonSCTLD_affected_Lower_winter",
      "SCTLD_RNA_Region_seasonSCTLD_affected_ECA_fall", 
      "SCTLD_RNA_Region_seasonSCTLD_affected_Lower_fall"),
    c("SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_fall", 
      "SCTLD_RNA_Region_seasonSCTLD_unaffected_Lower_fall",
      "SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_winter", 
      "SCTLD_RNA_Region_seasonSCTLD_unaffected_Lower_winter"),
listValues = c(0.25,0.25,0.25,0.25,-0.25,-0.25,-0.25,-0.25) )

resOrdered_affvsunaff<- as.data.frame(
               
                                 res_affvsunaff[order(res_affvsunaff$padj),])
                
DEG_affvsunaff <-resOrdered_both_season %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_both_season)) %>%
   
                                filter(padj<0.1)
```


```{r}
res_affvsunaff_ECA<- results(dds_sctld.reg.season, contrast = list(
    c("SCTLD_RNA_Region_seasonSCTLD_affected_ECA_winter", "SCTLD_RNA_Region_seasonSCTLD_affected_ECA_fall"),
    c("SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_winter", "SCTLD_RNA_Region_seasonSCTLD_unaffected_ECA_fall")
  ),
  listValues = c(0.5, -0.5) )

resOrdered_affvsunaff_ECA<- as.data.frame(
               
                                 res_affvsunaff_ECA[order(res_affvsunaff_ECA$padj),])
                
DEG_affvsunaff_ECA <-resOrdered_affvsunaff_ECA %>%  
  
                                mutate(ofav_gene = rownames(resOrdered_affvsunaff_ECA)) %>%
   
                                filter(padj<0.1)
```


```{r}

ids_affvsunaff<- DEG_affvsunaff_ECA$ofav_gene

count_affvsunaff_vst <- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% ids_affvsunaff),]

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_affvsunaff_vst))
all(rownames(colData_4M) == colnames(count_affvsunaff_vst))

```

# Combined contrast
```{r}
gene_ECA_winter<- DEG_affvsunaff_ECA$ofav_gene

# gene_ECA_winter<- gene_ECA_winter[order(gene_ECA_winter$log_ECA_winter, decreasing = TRUE),]

ECA_winter<- colData_4M %>% filter(Region == "ECA")
samples_ECA_winter<- ECA_winter$Sample_ID

samples_com.com.WvsF<-samples_ECA_winter
vst_out_ECA_winter<-count_affvsunaff_vst[(rownames(count_affvsunaff_vst) %in% gene_ECA_winter),]

vst_out_ECA_winter<-count_affvsunaff_vst[,(colnames(count_affvsunaff_vst) %in% samples_ECA_winter)]

samples_com.com.WvsF<-samples_ECA_winter
vst_out_com.com.WvsF<-vst_out_ECA_winter

# making the rownames and column names identical
all(rownames(samples_com.com.WvsF) %in% colnames(vst_out_com.com.WvsF))
all(rownames(samples_com.com.WvsF) == colnames(vst_out_com.com.WvsF))

mat_com.com.WvsF<-assay(vst_out_com.com.WvsF) #sig genes x samples
mat_com.com.WvsF.scaled <- t(apply(mat_com.com.WvsF, 1, scale)) #center and scale each column (Z-score) then transpose
colnames(mat_com.com.WvsF.scaled)<-colnames(mat_com.com.WvsF)

sctld.WvsF <- as.factor(ECA_winter$SCTLD_RNA)
sctld.WvsF <- factor(sctld.WvsF,levels = c("SCTLD_unaffected","SCTLD_affected"))

region.WvsF <- as.factor(ECA_winter$Region)
region.WvsF <- factor(region.WvsF,levels = c("ECA"))

SP.WvsF <- as.factor(ECA_winter$sample_period)
SP.WvsF <- factor(SP.WvsF,levels = c("SP2","SP3"))


ha_com.com.WvsF <- HeatmapAnnotation( 
                         Season  = SP.WvsF,
                         Region = region.WvsF,
                         SCTLD = sctld.WvsF,
                         col = list(
                           Season = c("SP2"= "#D55E00","SP3"="#0072B2"),
                           Region = c("ECA"="#009E73"),
                           SCTLD = c("SCTLD_unaffected" ="#56B4E9","SCTLD_affected" = "#E69F00")
                           ),
                         annotation_legend_param = list(
                           Season = list(title = "Season"),
                           Region = list(title = "Region"),
                           SCTLD = list(title = "SCTLD susceptibility")),
                           show_legend = c(Season = TRUE, Region = TRUE, SCTLD = TRUE) 
                         )
  # sample_data = anno_barplot(mat_com.com.WvsF, gp = gpar(fill = "blue"), height = unit(2, "cm"))  # Barplot for samples

```

```{r}
Heatmap(mat_com.com.WvsF.scaled, 
        cluster_columns =  T,  # Disable clustering for columns
        cluster_rows =F,     
        top_annotation = ha_com.com.WvsF)
```
































































































############ Extra code ##########
### Writing tables and data counts
#### vst transformed read counts with different DEG lists
```{r}
# write_rds(count_region_2seasons_vst, "cache/count_region_2seasons_vst.rds")

```


```{r}
genes_uniq_unaff<- uniq_unaff_month_region$ofav_gene

count_data_uniq_unaff<- vst_dds_sctld.reg.season[(rownames(vst_dds_sctld.reg.season) %in% genes_uniq_unaff),]

#head(sample_data_Lower)
#names(sample_data_Lower)

# making the rownames and column names identical
all(rownames(colData_4M) %in% colnames(count_data_uniq_unaff))
all(rownames(colData_4M) == colnames(count_data_uniq_unaff))

count_data_uniq_unaff_df<- assay(count_data_uniq_unaff)
count_data_uniq_unaff_df<- as.data.frame(count_data_uniq_unaff_df)
```

#### Writing vst transformed read counts with for genes with an adjspval of 0.5
```{r}

# write_rds(vst_0.5_filt_sctld.reg.season, "cache/vst_0.5_filt_sctld.reg.season.rds")

```



### filtering to write
```{r}

DEG_ECA_aff_w<- DEG_ECA_aff[,] %>% select(-gDNA,-mRNA, -"CDS-transcript",-Translation)

DEG_ECA_unaff_w<- DEG_ECA_unaff[,] %>% select(-gDNA,-mRNA, -"CDS-transcript",-Translation)

DEG_lower_aff_w<- DEG_lower_aff[,]  %>% select(-gDNA,-mRNA, -"CDS-transcript",-Translation)

DEG_lower_unaff_w<- DEG_lower_unaff[,]  %>% select(-gDNA,-mRNA, -"CDS-transcript",-Translation)

common_aff_ECA_lower_w<- common_aff_ECA_lower %>% select(-gDNA,-mRNA, -"CDS-transcript",-Translation) 

common_unaff_ECA_lower_w<- common_unaff_ECA_lower %>% select(-gDNA,-mRNA, -"CDS-transcript",-Translation) 

```

### Setting up table to be written out
```{r}
uniq_unaff_w<- uniq_unaff_month_region %>% select(ofav_gene)

uniq_unaff_w<- uniq_unaff_w %>% inner_join(region_WvsF_anno, by="ofav_gene") %>% mutate(DEG_unaffected = 1)

uniq_aff_w<- uniq_aff_month_region %>% select(ofav_gene)

uniq_aff_w<- uniq_aff_w %>% inner_join(region_WvsF_anno, by="ofav_gene") %>% mutate(DEG_affected = 1)
```

### Setting up table to be written out
```{r}
uniq_aff_w<- uniq_aff_month_region %>% select(ofav_gene)

uniq_aff_w<- uniq_aff_w %>% inner_join(region_WvsF_anno, by="ofav_gene") %>% mutate(DEG_affected = 1)

uniq_aff_w<- uniq_aff_month_region %>% select(ofav_gene)

uniq_aff_w<- uniq_aff_w %>% inner_join(region_WvsF_anno, by="ofav_gene") %>% mutate(DEG_affected = 1)

uniq_aff_w<- uniq_aff_w %>% select("ofav_gene","log_aff_ECA","padj_aff_ECA","baseMean_aff_ECA",
"log_unaff_ECA","padj_unaff_ECA","baseMean_unaff_ECA","log_aff_lower",
"padj_aff_lower","baseMean_aff_lower","log_unaff_lower","padj_unaff_lower","baseMean_unaff_lower","sig_genes.WvsF","sig_unaff.region_WvsF","sig_aff.region_WvsF","GeneID","TranscriptID","Feature","Contig","Start","Stop","Strand","Name","Product","Alias/Synonyms","EC_number","BUSCO","PFAM","InterPro","EggNog","COG","Goterms","Secreted","Membrane","Protease","CAZyme","Notes","DEG_affected","mRNA","CDS-transcript","Translation","GO_terms","KO","ann_scores")
```


#### Writing results tables
```{r}

# write_csv(DEG_ECA_aff_w, "results/5_All_4M_DEG/DEG_ECA_affected.csv")
# 
# write_csv(DEG_ECA_unaff_w, "results/5_All_4M_DEG/DEG_ECA_unaffected.csv")
# 
# write_csv(DEG_lower_aff_w, "results/5_All_4M_DEG/DEG_Lower_affected.csv")
# 
# write_csv(DEG_lower_unaff_w, "results/5_All_4M_DEG/DEG_Lower_unaffected.csv")
# 
# write_csv(uniq_unaff_w, "results/5_All_4M_DEG/DEG_uniq_unaff.csv")
# 
# write.csv(uniq_aff_w, "results/5_All_4M_DEG/DEG_uniq_aff.csv", row.names = FALSE, quote = TRUE)
# 
# write_csv(common_aff_ECA_lower_w,"results/5_All_4M_DEG/common_aff_ECA_Lower.csv")
# 
# write_csv(common_unaff_ECA_lower_w,"results/5_All_4M_DEG/common_unaff_ECA_Lower.csv")

```




```{r}
sessionInfo()
```
